---
title: "*Principal Component Analysis*"
author: "S.Tonini, F. Chiaromonte (special thanks to J. Di Iorio and L. Insolia)"
date: "February 16th 2023"
output:
  pdf_document:
    toc: true
  html_document: default
urlcolor: blue
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) # just to clean the pdf files, not recommended in practice!
```


\section{Introduction}

Working in high-dimensional spaces can be undesirable for several reasons so it could be useful to project the features to a space of fewer dimensions.

Principal Component Analysis (PCA) is a \textit{Dimension Reduction} technique for unsupervised problems. 
It projects/transforms the data into a lower-dimensional space so that the lower-dimensional representation retains some meaningful properties of the original data, ideally close to its intrinsic dimension.

\subsection{Libraries}
We are going to use:
\begin{itemize}
  \item \textbf{mvtnorm}: \textit{Multivariate Normal and t Distributions}
  \item \textbf{factoextra}: \textit{Extract and Visualize the Results of Multivariate Data Analyses}
  \item \textbf{scales}: \textit{Scale Functions for Visualization}
  \item \textbf{ellipse}: \textit{Functions for drawing ellipses and ellipse-like confidence regions}
  \item \textbf{corrplot}: \textit{Visualization of a Correlation Matrix}
\end{itemize}

```{r}
library(mvtnorm)    # for the toy simulated example
library(factoextra) # contains also decathlon data
library(scales)     # to create ggplot-like figures in base R (color transparency)
library(ellipse)    # to add elliptical confidence regions in base R plots
library(corrplot)   # correlation plots
```

\subsection{Data} 
Today we are going to use the \textbf{decathlon2}  data set, which is available in the \textbf{factoextra} package.
It consists of 27 observations (athletes) and the following 13 variables (performance).
```{r}
help(decathlon2)
head(decathlon2)
```


\section{A Toy Example of Dimension Reduction}

Let's simulate a 2-dimensional data set from a Gaussian distribution with higher dispersion along the $y$-axis.

```{r}
set.seed(1234)                # set seed for reproducibility
mu  <- c(1, 2)                # location/mean vector (p times 1)
sig <- cbind(c(1,1), c(1,4))  # covariance matrix (p times p)
n   <- 100                    # number of points

X <- rmvnorm(n, mu, sig)      # data generation
colnames(X) <- c("x","y")     # renaming
head(X)                       # visualize our data
plot(X, asp=1)                # plot our data (with fixed aspect ratio)

points(mu[1], mu[2],          # add centroid
       col=alpha("red", 0.5), # marker color and transparency
       pch=3,                 # marker type
       lwd=5,                 # marker width
       cex=2)                 # marker size

# iterate across confidence levels
conflev <- c(0.5, 0.75, 0.9, 0.95, 0.99, 0.999)
for (confi in conflev){
  lines(ellipse(sig, centre=mu, level=confi),    # add elliptical confidence regions
        col = alpha("red", 0.5),                 # add color and transparency
        lty = 3)                                 # dashed lines
}

```

How can we reduce the dimension of these data? 

We can reduce the number of features (columns) in different ways.
Let's see some.

\subsection{Using the sample mean}
The sample mean is the easiest 0-dimensional data reduction method, because it allows to collapse all the information into a single point.
```{r}
med <- colMeans(X)              # column-wise means
med

plot(X, asp=1,                  # plotting the points
     col=alpha('black', 0.5))   # reduce the transparency

points(med[1], med[2],          # plotting their centroid
       col='red', pch=3,        # color, marker type
       cex = 2, lwd=3)          # size, width
```

What is the variance?

What is the associated loss of information?

What is the error?

We can think of the error in this way: 
```{r}

plot(X, asp=1,                  # plotting the points
     col=alpha('black', 0.5))   # reduce the transparency   

for(i in 1:n){                  # for each point
  lines(rbind(X[i,], med),      # plot the "error"
        col=alpha('blue',0.2))
}  

points(med[1], med[2],          # plotting their centroid
       col='red', pch=3,        # color, marker type
       cex = 2, lwd=3)          # size, width


```

We are collapsing our data to one point, the sample mean (also known as PC0). The error is high.

\subsection{Projecting the data on a new axis}
We can easily identify two axes from the sample mean: an $horizontal$-axis, and a $vertical$-axis.
```{r}

plot(X, asp=1, col=alpha('black', 0.5))     # data

abline(h=med[2], lty=2)                     # horizontal axis 
points(X[,1], rep(med[2], n),               # projected points
       col=alpha('red',0.5))

abline(v=med[1], lty=2)                     # vertical axis
points(rep(med[1], n), X[,2],               # projected points
       col=alpha('blue',0.5))   

for(i in 1:n){                  
  lines(rbind(X[i,], c(X[i,1],med[2])),     # plot the horiz. proj. "error"
        col=alpha('red',0.1))
  
  lines(rbind(X[i,], c(med[1], X[i,2])),    # plot the vert. proj. "error"
    col=alpha('blue',0.1))
}  

points(med[1], med[2], col='green',         # centroid
       pch=3, cex = 1, lwd=3)

legend(-10,6,                               # legend location
       c("horiz. proj.","vert. proj","mean"),
       col=c("red","blue","green"), 
       pch=c(1,1,3))

# comparing variances
var(X[,1])                                  # red dots variance
var(X[,2])                                  # blue dots variance
```

Which of the two axis maximizes the variance?
```{r}
var(X[,2]) > var(X[,1])
```

The vertical axis. 
Blue points are more scattered and the error (sum of the lenghts of blue segments) is lower.

```{r}
# 2 times 1 panels in a figure
par(mfrow=c(1,2))                           

# left panel: horiz. axis
plot(X, asp=1, col=alpha('black', 0.5))
abline(h=med[2], lty=2)
points(X[,1], rep(med[2], n), col='red')
for(i in 1:n)
  lines(rbind(X[i,], c(X[i,1], med[2])), col=alpha('red',0.3))

# right panel: vert. axis
plot(X, asp=1, col=alpha('black', 0.5))
abline(v=med[1], lty=2)
points(rep(med[1], n), X[,2], col='blue') 
for(i in 1:n)
  lines(rbind(c(med[1],X[i,2]),X[i,]), col=alpha('blue',0.3))

# reset it as 1 x 1
par(mfrow=c(1,1))                           
```

Using this strategy we can find the "best" axis -- i.e., the one maximizing the variance.
Let's try a brute force approach.
```{r}

# Compute the variance across (almost) all possible directions
theta   <- seq(0, 2*pi, by = 2*pi/360)    # angles
Var     <- rep(NA, length(theta))

# centered data
Xc = scale(X, scale=F)

for(i in 1:length(theta))
{
  a      <- c(cos(theta[i]), sin(theta[i]))  # directional vector
  proj   <- Xc %*% a                         # projecting points
  Var[i] <- var(proj)                        # computing variance
}

# plotting the variance for each direction/angle
plot(theta, Var, type = 'l', 
     col='red', lwd = 2)
abline(v=c(0, pi/2, pi, 3/2*pi, 2*pi),       # fundamental angles
       lty=2)    
```


The direction with highest variability is identified by the maximun of the function.
```{r}
max.var   <- max(Var) # maximum variance
# max.theta <- theta[which.max(Var)] # theta angle with maximum variance (only the 1st)
max.theta <- theta[Var==max.var] # theta angle with maximum variance (not only the 1st)
max.theta

# projected data (see that their variance is equal to the maximum)
projx <- X %*% c(cos(max.theta[1]), sin(max.theta[1]))
var(projx)
max.var

# plotting the Variance for each direction/angle
plot(theta, Var, type = 'l', col='red', lwd = 2, lty=1)
abline(v=c(0, pi/2, pi, 3/2*pi, 2*pi), lty=2) # fundamental angles
points(max.theta, rep(max.var, length(max.theta)), 
       pch=4, col='blue', lwd=3, cex=2)
```

We just found the first principal component (PC1) as the axis maximizing the variance.
Let's plot the solution we have found.

\textit{Remark}: there is rotational invariance, so we can pick any maximizer theta (e.g., the first one).
```{r}

# slope
slopepc1 <- sin(max.theta[1])/cos(max.theta[1])

# plot centered data
plot(Xc, asp=1, col='black')
abline(v=0, lty=3)
abline(h=0, lty=3)
# plot PC1
abline(a=0,b=slopepc1, col = "red", lty=1, lwd=2)

```

Note  that there is only another component orthogonal to PC1 (and centered at 0) since we are in a 2-dimensional setting.

Thus, we can also view PCA as a change of bases/rotation.
```{r}

# rotation matrix
R1 <- c(cos(max.theta[1]), -sin(max.theta[1]))
R2 <- c(sin(max.theta[1]), cos(max.theta[1]))
R <- rbind(R1, R2)

# plot rotated data such that PC1 and PC2 are now the "standard" bases
plot(Xc %*% R, asp=1)
abline(v=0, col="red",lty=2, lwd=2)
abline(h=0, col="red",lty=2, lwd=2)

```


Finally, notice that maximizing the variance is equivalent to minimizing squared distances.
```{r}

# Compute the squared distances across (almost) all possible directions
theta2   <- seq(0, 2*pi, by = 2*pi/360) # angles
Dist    <- rep(NA, length(theta))

for(i in 1:length(theta))               # for each angle theta
{
  mi = sin(theta2[i])/cos(theta2[i])    # slope of the associated line
  distk = rep(NA, n)                    # initialize point-to-line distances
  for (k in 1:n){                       # compute them
    distk[k] = abs(-mi*Xc[k,1]+Xc[k,2])/sqrt(mi^2+1)
  }
  Dist[i] = mean(distk^2)               # store average of squared distances
}

# min and argmin for the distance method
min.dist   <- min(Dist)                            # min distance
min.theta2 <- theta2[Dist==min.dist]               # theta angle with min distances

# compare the two approaches

# plotting the Variance for each direction/angle
plot(theta, Var, type = 'l', col='red', 
     lwd = 2, lty=1, ylab = "Objective value")
abline(v=c(0, pi/2, pi, 3/2*pi, 2*pi), lty=2)      # fundamental angles
points(max.theta, rep(max.var, length(max.theta)), 
       pch=4, col='blue', lwd=3, cex=2)

# plotting the avg squared distances for each direction/angle
points(theta2, Dist, type = 'l', 
       col='cyan', lwd = 2, lty=1)
abline(v=c(0, pi/2, pi, 3/2*pi, 2*pi), lty=2)      # fundamental angles
points(min.theta2, rep(min.dist, length(min.theta2)),
       pch=4, col='blue', lwd=3, cex=2)

legend(4,3,
       c("Variance","Avg. sq. distances"),
       col=c("red","cyan"), 
       lty=c(1,1),
       bg="white")
```


Global minima/maxima do differ, but are the arguments (theta values) minimizing/maximizing these functions the same?
```{r}
max.theta
min.theta2
```

The same result can be obtained through low-rank approximations.

The functions \textbf{princomp()} and \textbf{prcomp()} in the \textbf{stats} package use the spectral decomposition and the singular value decomposition (SVD), respectively.
We will focus on the latter.


\section{PCA on decathlon2 data}

We are going to perform PCA to the first 10 columns of the \textbf{decathlon2} data set to analyze athletes performance.

```{r}
# select only the first 10 columns
decathlon2<- decathlon2[, 1:10]
head(decathlon2)
dim(decathlon2)
summary(decathlon2)
```

To perform PCA we use the function \textbf{prcomp}.

```{r}
help(prcomp)
```

Among other things, the function takes as input:
\begin{itemize}
  \item \textbf{data:} a data frame
  \item \textbf{scale:} a logical value (TRUE/FALSE) indicating whether the variables should be scaled to have unit variance before the analysis takes place
\end{itemize}
Note that by default variables are centered to have zero mean.

Therefore, to perform PCA after scaling the data, we run:

```{r}
res <- prcomp(decathlon2, scale = TRUE)
str(res)
```

The result is a list containing 5 elements:
\begin{itemize}
  \item \textbf{sdev:} the standard deviations of the principal components (square root of the eigenvalues)
  \item \textbf{rotation:} the matrix of variable loadings (eigenvectors)
  \item \textbf{center:} the centering used if \textbf{scale=TRUE}
  \item \textbf{scale:} the scaling used if \textbf{scale=TRUE}
  \item \textbf{x:} the scores, i.e. the rotated data
\end{itemize}

\section{Selecting the number of components}
The selection of a "suitable" number of components is non-trivial since in unsupervised learning tasks no ground-truth/labels/response variable is available (which may be used to assess and validate our results otherwise, e.g. through cross-validation -- more on this later). 

However, some "rules of thumb" have been developed.
For instance, one may consider the \textbf{percentage of variance explained} (\textbf{PVE}), or the \textbf{cumulative PVE}.

Here we use the \verb|get_eig| function of the \textbf{factoextra} package.
```{r}
help(get_eig)
# eigenvalue, PVE and cumulative PVE for each PC
get_eig(res)
```


Based on this information we can create a \textbf{scree plot}, and try to find an \textit{elbow} (i.e. an inflection point) therein.
```{r}

fviz_eig(res, addlabels = TRUE, ylim = c(0, 50))

```

Sometimes, it is not easy to identify an elbow, or this might be associated to a particularly low cumulative PVE.

Thus, it is common to fix an \textbf{acceptance threshold} a priori (e.g. 80%), and then look at the \textbf{cumulative PVE}.

```{r}
plot(get_eig(res)$cumulative.variance.percent, 
     type='b', axes=F, xlab='Dimensions', ylab='cumulative PVE', ylim=c(0,100))
abline(h=100, col=alpha('blue',0.5))
abline(h=80, lty=2, col='red', lwd=2) # thresholding
box()
axis(2, at=0:100,labels=0:100)
axis(1,at=1:ncol(decathlon2),labels=1:ncol(decathlon2))
grid()
```

\section{Loadings interpretation}
Let's focus on the loadings, i.e., the eigenvectors representing the directions of the PCs.

```{r}
loadings <- res$rotation
loadings
```

We can plot the first two PCs (PC1 and PC2) in the \textbf{graph of variables} (also called correlation circle).

Here the importance of the original features is represented by the \textbf{color code}:
\begin{itemize}
  \item red: \textit{high}
  \item blue: \textit{medium} 
  \item white: \textit{low}
\end{itemize}
and by the lenght of the vector (its \textit{closeness} to the circle).

In these plots, positively correlated variables have same direction (they are close to each other, i.e. they have a small angle in between).
Negatively correlated variables have opposite directions.
Uncorrelated variables are orthogonal.

\textit{Remark:} a lower value for "X100m", "X400m", "X110m hurdle" means a better performance. 
So it makes sense that they are negatively correlated with "long jump".

```{r}
fviz_pca_var(res,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE )    # Avoid text overlapping

```

We also see that "X100m", "Long jump" and "High jump" and "Pole vault" contribute the most to PC1 and PC2.

If we want to show also the individuals we can run the \textbf{biplot of individuals and variables}.
Here similar athletes are grouped together.

```{r}
fviz_pca_biplot(res, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
)
```

Along PC1 we tend to find athletes with higher/lower overall performance, which are mainly related to agility.
On PC2 we can distinguish stronger athletes from others.

\textit{Remark:}  observations are represented by their projections, but variables are represented by their correlations.

Information regarding all the PCs (e.g., the first 4) can be obtained in the following way:
```{r}
plot.new()
par(mar = c(1,4,0,2), mfrow = c(4,1))
for(i in 1:4)
{
  barplot(loadings[,i], ylim = c(-1, 1),
          ylab=paste0("PC", i))           # this is not the axis label, but the plot title
  abline(h=0)
}
```


The "importance" of each variable across different PCs is also contained in the field \textbf{cos2} of the function \verb|get_pca_var|:
\begin{itemize}
  \item A high cos2 indicates a good representation of the variable on the principal component. In this case the variable is positioned close to the circumference of the correlation circle.
  \item A low cos2 indicates that the variable is not perfectly represented by the PCs. In this case the variable is close to the center of the circle.
\end{itemize}
For a given variable, the sum of the cos2 on all the principal components is equal to one.

```{r}
help(get_pca_var)
varpca <- get_pca_var(res)
varpca

corrplot(varpca$cos2, is.corr=FALSE)
```


\section{Robust PCA}

Unlike classical PCA, there exist robust versions of PCA (e.g., ROBPCA) that are resistant to outliers in the data. 

Here robust loadings are computed in a robust manner (e.g., through projection-pursuit techniques and the minimum covariance determinant method) to remove the effects of outliers.

We will focus on the \textbf{PcaHubert} function within the \textbf{rrcov} package, and itdepends on a trimming proportion $\alpha=h/n$ denoting the fraction of points (with larger distances) that do not contribute to the objective function (in this case $h$ out of $n$).

```{r}
## 

set.seed(123)
library(rrcov)
help(PcaHubert)

# PCA of the Hawkins Bradu Kass's Artificial Data
# The first 14 observations are outliers, created in two groups: 
# 1--10 (more difficult to detect) and 11--14.
data(hbk)
pairs(hbk)

pcaRob <- PcaHubert(hbk, k=4, alpha=0.75)
print("Robust output")
pcaRob
pcaRob$loadings

## Compare with the classical PCA
pca <- prcomp(hbk)
cat("\n\n")               # line break
print("Classical output")
pca

## Compare with the classical PCA on clean data
pcaclean <- prcomp(hbk[15:nrow(hbk), ])
cat("\n\n")               # line break
print("Classical output on clean data")
pcaclean


```

Let's plot their solutions:
```{r}

par(mfrow=c(1,2)) 

plot(pcaRob)                    # distance plot

pca2 <- PcaHubert(hbk, k=2)
plot(pca2)                      # PCA diagnostic plot (or outlier map)
```


```{r}

par(mfrow=c(1,3)) 

## Use the standard plots available for prcomp and princomp
screeplot(pcaRob, main="Robust PCA")
screeplot(pca ,main="PCA")
screeplot(pcaclean ,main=" clean PCA")
```


```{r}
par(mfrow=c(1,3)) 

biplot(pcaRob)
title("Robust PCA", line = 3)

biplot(pca)
title("PCA", line = 3)

biplot(pcaclean)
title("clean PCA", line = 3)


```

Retrieve the robust covariance matrix:
```{r}

print("Cov by Robust PCA (25% trimming)")
py <- PcaHubert(hbk, alpha=0.75)
cov.1 <- py@loadings %*% diag(py@eigenvalues) %*% t(py@loadings)
cov.1    

cat("\n\n")
print("Cov by another robust method")
CovRobust(hbk, CovControlSest(method="auto",))

cat("\n\n")
print("Cov on clean data")
cov(hbk[15:nrow(hbk), ])

cat("\n\n")
print("Empirical output (i.e. 0% trimming)")
cov(hbk)

```




\section{Multidimensional Scaling}
Multidimensional scaling (MDS) is useful to visualize the level of similarity of individual cases in a dataset. 
MDS is used to translate "information about the pairwise 'distances' among a set of "n" objects or individuals" into a configuration of "n" points mapped into an abstract Cartesian space.

See for instance:
\begin{itemize}
  \item cmdscale() [stats package]: Classical (metric) multidimensional scaling.
  \item isoMDS() [MASS package]: Kruskal’s non-metric multidimensional scaling (one form of non-metric MDS).
  \item sammon() [MASS package]: Sammon’s non-linear mapping (one form of non-metric MDS).
\end{itemize}

These functions require a distance object as input and the number of dimensions in the reduced space (by default equal to 2).
