---
title: "How to approach data"
author: "SSSA - Applied Statistics - Chiara Seghieri and Costanza Tortù"
date: "2023-11-07"
output:
  pdf_document: default
  html_document: default
---

## Preliminaries

### Recall packages

Before loading packages, please install it! To install a package called "package" you can use the function install.packages("package")

```{r setup, include=FALSE}

options(warn=-1)


library(VIM) # Useful to analyze missing data  
library(mice) # Useful to analyze missing data 
library(foreign) # Package that allows you to read .dta data


```

### Import Data

```{r}

crime <- read.dta("~/Documents/Sant'Anna/Corso allievi/Data/crimedata.dta")


```


## Have a first look at data


### Inspect variables


```{r}
dim(crime) # units x variables
colnames(crime) # have a look at the names of the variables
```

#### Look at the joint distribution of sex and having experienced at least one crime in the previous year


```{r}

table(crime$bcsvictim, crime$sex)


```


### Look at the distribution of the two continuous scores

Note: the na.omit command is used to exclude missing data from the computation.

worryx: Worry about being a victim of crime (high score = high level of worry) (Module C)



```{r}


hist(na.omit(crime$worryx),
     main = "Worry about crime", 
     col = "red")

```
antisocx: Anti-social behaviour in their neighbourhood (high score =high levels of anti-social behaviour)



```{r}


hist(na.omit(crime$worryx), 
     main = "Anti-social behaviour", 
     col = "springgreen")

```

## Analyze missing data

There are a lot of missing data!!!!

### Count missings 

```{r}

colSums(is.na(crime)) # Look at the number of missings in each variable
table(complete.cases(crime)) # Look atthe number of individuals who have no missings 


```


### Look at patterns of missingness 

It allows you to inspect which combinations of variables are likely to be jointly missing


```{r}
MP_plot_crimedata <- aggr(crime[, which(colSums(is.na(crime)) > 1000)], 
                          col=c('red','blue'),
                          numbers=TRUE, sortVars=TRUE,
                          labels=names(crime), cex.axis=.7,
                          gap=3, ylab=c("Proportion of Missing Data","Pattern"))

```