---
title: "Confidence intervals"
author: "SISS - Applied Statistics - Chiara Seghieri and Costanza Tortù"
date: "2023-11-13"
output:
  pdf_document: default
  html_document: default
---

## Preliminaries

### Recall packages

```{r setup, include=FALSE}

options(warn=-1)
rm(list=ls())

library(VIM) # Useful to analyze missing data 
library(mice) # Useful to analyze missing data 
library(foreign) # Package that allows you to read .dta data
library(psych)
library(MatchIt)
library(bestNormalize)
library(corrplot)
library(GGally)
library(haven)
library(dataset)

```

### Import Data



```{r}

rm(list=ls())
value <- read_dta("~/Documents/Sant'Anna/Corso allievi/Data/Value Survey descrittive CI e test/WV6_Data_stata_vold.dta")


```


## Have a first look at data

```{r}

dim(value)
colnames(value)
head(value)


```

### Simplify data 

Here we apply some simplifications on data 
- i) Here we keep only observations with no missings (this is not the right procedure to deal with missings of course :-))
- ii) We focus on a subsample of countries (#360 Romania 255 Germany 380 Sweden 230 Spain)

```{r}
value <- value[complete.cases(value),]

included_countries <- c(360,255,380,230)
value <- value[which(value$cow %in% included_countries),]
value$cow <- as.factor(value$cow)
levels(value$cow) <- c("Spain", "Germany", "Romania", "Sweeden")


```


### Inspect variables


```{r}


quantitative_variables <- c("lifesat", "age", "relativism", "scepticism", "equality", "choice","voice")
dummies <- c( "male", "trust")
factors <- c("cow", "education")
qualitative_variables <- c(dummies, factors)




```


### Look at the distribution of the equality index

```{r}

hist(value$equality, 
     main ="Histogram of Equality index",
     col = "red",
     xlab = "Equality index")

summary(value$equality)

boxplot(value$equality,
        col="purple")
```

Let's have a look at the correlation among continuous variables 

```{r}

correlation_matrix <- cor(value[, c(quantitative_variables)])

corrplot(correlation_matrix, type = "upper", 
         order = "hclust", 
         tl.col = "black", tl.srt = 45)


```



## Compute confidence intervals for the average Equality index


### Compute sample mean

```{r}

  sample.mean <- mean(value$equality)
  print(sample.mean)

```



### Compute sample variance

```{r}

sample.n <- length(value$equality)
sample.sd <- sd(value$equality)
sample.se <- sample.sd/sqrt(sample.n)
print(sample.se)

```


### Find the t-score


```{r}
alpha = 0.05
degrees.freedom = sample.n - 1
t.score = qt(p=alpha/2, df=degrees.freedom,lower.tail=F)
print(t.score)
```

### Compute margin of error

```{r}
margin.error <- t.score * sample.se
print(margin.error)

```


### Now we are ready to compute the confidence interval 

```{r}
lower.bound <- sample.mean - margin.error
upper.bound <- sample.mean + margin.error
print(c(lower.bound,upper.bound))

```


## Compute confidence intervals for the average Equality index in a subsample


### Compute sample mean

```{r}
  
  N_subsample <- 250
  sampled_units <- sample(value$ID, N_subsample, replace = F)
  value_subsample <- value[which(value$ID %in% sampled_units),]
  sample.mean <- mean(value_subsample$equality)
  print(sample.mean)

```



### Compute sample variance

```{r}

sample.n <- length(value_subsample$equality)
sample.sd <- sd(value_subsample$equality)
sample.se <- sample.sd/sqrt(sample.n)
print(sample.se)

```


### Find the t-score


```{r}
alpha = 0.05
degrees.freedom = sample.n - 1
t.score = qt(p=alpha/2, df=degrees.freedom,lower.tail=F)
print(t.score)
```

### Compute margin of error

```{r}
margin.error <- t.score * sample.se
print(margin.error)

```


### Now we are ready to compute the confidence interval 

```{r}
lower.bound <- sample.mean - margin.error
upper.bound <- sample.mean + margin.error
print(c(lower.bound,upper.bound))

```


## Compute confidence intervals for the variance of Equality index 




### Compute sample variance

```{r}

sample.n <- length(value$equality)
sample.var <- var(value$equality)
print(sample.var)

```


### Find the chi-scores


```{r}
alpha = 0.05
degrees.freedom = sample.n - 1
chi.scores = qchisq(c(1-alpha/2, alpha/2), df =  degrees.freedom)
print(chi.scores)


```
### Now we are ready to compute the confidence interval 

```{r}
lower.bound <- degrees.freedom*sample.var/chi.scores[1]
upper.bound <- degrees.freedom*sample.var/chi.scores[2]

print(c(lower.bound,upper.bound))

```
## Confidence intervals for the difference in means

Let's focus on two countries: Germany and Romania. We want to build up a confidence interval for the difference im the mean equality index in the two countries.

1) Compute all the quantities you need 

```{r}
n1 <- length(which(value$cow == "Germany"))
xbar1 <- mean(value$equality[which(value$cow == "Germany")])
s1 <- var(value$equality[which(value$cow == "Germany")])
n2 <- length(which(value$cow == "Romania"))
xbar2 <- mean(value$equality[which(value$cow == "Romania")])
s2 <-  var(value$equality[which(value$cow == "Romania")])
```

2) Compute pooled variance

```{r}
sp = ((n1-1)*s1^2+(n2-1)*s2^2)/(n1+n2-2)
sp
```
3) Compute the margin of error

```{r}
margin <- qt(1-alpha/2,df=n1+n2-1)*sqrt(sp/n1 + sp/n2)
margin
```

4) Compute the confidence interval

```{r}
lowerinterval <- (xbar1-xbar2) - margin
lowerinterval

upperinterval <- (xbar1-xbar2) + margin
upperinterval
```

## Confidence intervals for the difference in proportions

Let's focus on two countries: Germany and Romania. We want to build up a confidence interval for the difference im the proportion of people who trust in people in the two countries.

1) Compute all the quantities you need 

```{r}
n1 <- length(which(value$cow == "Germany"))
p1 <- mean(value$trust[which(value$cow == "Germany")])

n2 <- length(which(value$cow == "Romania"))
p2 <- mean(value$trust[which(value$cow == "Romania")])

```

2) Compute the margin of error

```{r}
margin <- qnorm(1-alpha/2)*sqrt(p1*(1-p1)/n1 + p2*(1-p2)/n2)
margin
```

3) Compute the confidence interval

```{r}
lowerinterval <- (p1-p2) - margin
lowerinterval

upperinterval <- (p1-p2) + margin
upperinterval
```




