---
title: "Linear Regression - Example"
author: "SSSA - Applied Statistics - Chiara Seghieri and Costanza Tortù"
date: "2023-12-07"
output:
  pdf_document: default
  html_document: default
---

## Preliminaries

### Recall packages

```{r setup, include=FALSE}

options(warn=-1)
rm(list=ls())

library(VIM) # Useful to analyze missing data 
library(mice) # Useful to analyze missing data 
library(foreign) # Package that allows you to read .dta data
library(psych)
library(MatchIt)
library(bestNormalize)
library(corrplot)
library(GGally)

```

### Import Data

The data consists of a number of demographic variables (age, race, academic background, and previous real earnings), as well as a treatment indicator, and the real earnings in the year 1978 (the response).

Robert Lalonde, "Evaluating the Econometric Evaluations of Training Programs", American Economic Review, Vol. 76, pp. 604-620

```{r}
rm(list=ls())
data("lalonde")


```


## Have a first look at data


```{r}
dim(lalonde)# units x variables
head(lalonde)


```


### Inspect variables


```{r}

colnames(lalonde)

quantitative_variables <- c("age", "educ", "re74", "re75", "re78")
qualitative_variables <- c("treat", "race", "married", "nodegree")
dummies <- c("treat", "married", "nodegree" )

lalonde$treat_factor <- as.factor(lalonde$treat)
lalonde$married_factor <- as.factor(lalonde$treat)
lalonde$nodegree_factor <- as.factor(lalonde$nodegree)

all_variables <- c(quantitative_variables, qualitative_variables_factors)





```

Let's focus on quantitative variables

```{r}

pairs.panels(lalonde[, quantitative_variables], 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )

correlation_matrix <- cor(lalonde[, quantitative_variables])

corrplot(correlation_matrix, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)

```
You may also treat dummies as quantitative variables and compute correlation, but pay attention to the interpretation!!!!!!


```{r}

correlation_matrix_withdummies <- cor(lalonde[, c(quantitative_variables,dummies)])

corrplot(correlation_matrix_withdummies, type = "upper", 
         order = "hclust", 
         tl.col = "black", tl.srt = 45)

```

```{r}

ggpairs(lalonde[, all_variables],
        progress = F,
        lower=list(combo=wrap("facethist",  
        binwidth=0.5)))

```





# Run a regression model

The response variable measures earnings in 1978 while the marital status the age, the education, the race and the training program are independent variables.

## Make sure your data meet the normality assumption 

Let's have a look at the distribution of earnings in 1978

```{r}

hist(lalonde$re78, 
     main ="Histogram of real earnings in 1978",
     col = "red",
     xlab = "Earnings in 1978")



```
This is far from normality, let's apply a normalization trasformation

```{r}

re78_BN <- bestNormalize(lalonde$re78)
re78_BN

lalonde$re78_normalized <- re78_BN$x.t

hist(lalonde$re78_normalized,
     main ="Histogram of normalized real earnings in 1978",
     col = "orange",
     xlab = "Normalized Earnings in 1978")


```



## Simple Linear Regression

We investigate the relationship between the education and the real earnings in the year 1978

educ: years of education
re78: earnings in 1978


```{r}


simple_model <- lm(re78_normalized ~ educ, 
                   data = lalonde)

summary(simple_model)

plot(lalonde$age, lalonde$re78, pch=16)
abline(simple_model, col="red" )




```

Further inspect your model

```{r}

plot(simple_model)

```



## Multiple Regression

We investigate the determinants of real earnings in the year 1978

age:  gae (years), numeric
treat: attendence of a training program
educ: years of education
married: marital status
re78: earnings in 1978

```{r}


multiple_model_normalized <-lm(re78_normalized ~ age + educ + as.factor(race) 
                             + married + treat + nodegree, 
                               data = lalonde)

summary(multiple_model_normalized)


```
Furtehr inspect your model

```{r}
plot(multiple_model_normalized)

```



Add valuable interactions

```{r}


multiple_model_normalized_winteractions <-lm(re78_normalized ~ age + educ + as.factor(race) +
                                             married + treat + nodegree +
                                             as.factor(race)*educ + treat*married, 
                                             data = lalonde)

summary(multiple_model_normalized_winteractions)

```

