---
title: "Models for binary outcomes"
author: "SSSA - Applied Statistics - Chiara Seghieri and Costanza Tortù"
date: "2023-10-22"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Load data 

```{r}
library(readxl)
library(foreign)
library(haven)
titanic <- as.data.frame(read_excel("~/Documents/Sant'Anna/Corso allievi/Data/TITANIC/TITANIC.xlsx"))
mytitanic<-titanic[,c("survived","pclass","age","Gender","cabin","sibsp","parch")]



```

# Have a preliminary overview of data

## look  at the columns 

```{r}
ncol(titanic)
head(mytitanic)

```

## Are there missing entries?


```{r}
summary(titanic)

```


## Inspect variables

```{r}

my_cols <- c( "red","springgreen")
pairs(mytitanic[,c("survived","pclass","age","Gender","sibsp","parch")], 
      pch = 19,  cex = 0.5,  col = my_cols[mytitanic$survived + 1],
      lower.panel=NULL)

```

# Inspect the outcome of interest 

```{r}
table(mytitanic$survived)
hist(mytitanic$survived, main = "Survived" )

```
# Fit your models

First, we investigate the relationship between age and survival 

```{r}
boxplot(age ~ survived,data = mytitanic, 
        col = "bisque",
        xlab = " Survival",
        ylab = "Age (in years)",
        main = "Incidence of survival vs. Age")
```

```{r}
cdplot(as.factor(survived)~age,mytitanic,
       col=c("lightgoldenrod", "lightcyan"), 
       ylab = "Survival", 
       xlab ="Age", main = "Conditional density plot")
```


## Compare the approaches for binary data with a naive linear approach


### Naive linear model 

```{r}


simple_naive_lm <- lm(survived ~ age,
              data = mytitanic)
summary(simple_naive_lm)

simple_naive_lm_predicted <- predict(simple_naive_lm)

```

### Simple Logit Model 

```{r}

simple_logit_fit <- glm(survived ~ age, family="binomial"(link="logit"),
                    data = mytitanic)
summary(simple_logit_fit)
simple_logit_fit_predicted <- predict(simple_logit_fit,type="response")


```


```{r}

par(mfrow = c(1,2))

plot(mytitanic$age[which(!is.na(mytitanic$age))], main = "Linear fit", 
     simple_naive_lm_predicted, xlab = "Age", ylab = "Survived",
     pch=16, col="black", ylim=c(0,1) )

plot(mytitanic$age[which(!is.na(mytitanic$age))], main ="Logit fit",
     simple_logit_fit_predicted, xlab = "Age", ylab = "Survived",
     pch=16, col="red", ylim=c(0,1) )


```
Here you do not see relevant differences but in some cases LPM can predict probabilities that are out of the range [0,1]!!!

### Add a quadratic term for age

```{r}
 
simple_logit_fit2 <- glm(survived ~ age + I(age^2), family="binomial"(link="logit"),
                    data = mytitanic)
summary(simple_logit_fit2)
simple_logit_fit_predicted2 <- predict(simple_logit_fit2,type="response")

```

Compare the two models

```{r}
anova(simple_logit_fit,simple_logit_fit2,test='LR')
```


## Add covariates 



## Add other meaningful covariates 

```{r}

logit_fit <- glm(survived ~ age + I(age^2) + as.factor(pclass) +
                as.factor(Gender) +  sibsp + as.factor(parch), 
                family = "binomial"(link="logit"), 
                data = mytitanic)
summary(logit_fit)


```




## Add interactions

```{r}

logit_fit_winteract <- glm(survived ~ age + I(age^2) + as.factor(pclass) +
                as.factor(Gender) +  sibsp +  as.factor(parch)+ age*as.factor(Gender) +
                + as.factor(pclass)*as.factor(Gender), 
                family = "binomial", 
                data = mytitanic)
summary(logit_fit_winteract)


```


## The probit link

```{r}

probit_fit <- glm(survived ~ age + as.factor(pclass) +
                as.factor(Gender) +  sibsp + as.factor(parch), 
                family = binomial(link = "probit"),
                data = mytitanic)
summary(probit_fit)


```
```{r}

deviance(probit_fit)

deviance(logit_fit)

d <- deviance(probit_fit) - deviance(logit_fit)
d

```




