---
title: "Models for count data"
author: "SISS - Applied Statistics - Chiara Seghieri and Costanza Tortù"
date: "2023-12-04"
output:
  pdf_document: default
  html_document: default
---

# Load data 

```{r}
rm(list=ls())

library(VIM) # Useful to analyze missing data 
library(mice) # Useful to analyze missing data 
library(foreign) # Package that allows you to read .dta data
library(psych)
library(MatchIt)
library(bestNormalize)
library(corrplot)
library(GGally)
library(MASS)
library(AER)
library(pscl)

```

# Have a preliminary overview of data

## look  at the columns 

```{r}

data("DoctorVisits")
doc <- DoctorVisits

```

Data come from an Australian health survey, where visits is the number of doctor visits in past two weeks.

## Overview

```{r}
summary(doc)
head(doc)
```

## Preliminary inspection of the outcome variable

```{r}

counts <- table(doc$visits)
barplot(counts)


```

## Correlation in the data 

```{r}

quantitative_variables <- c("visits","age","income",
                            "illness","reduced","health")

pairs.panels(doc[,quantitative_variables], 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )

correlation_matrix <- cor(doc[, quantitative_variables])

corrplot(correlation_matrix, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```

# Fit your models

## Simple case 

### Naive linear model

```{r}

simple_naive_lm <- lm(visits ~ age,
              data = doc)
summary(simple_naive_lm)
simple_naive_lm_predicted <- predict(simple_naive_lm,
                                     type="response")

```
### Poisson model 

```{r}

simple_poisson <- glm(visits ~ age, family="poisson",
              data = doc)
summary(simple_poisson)
simple_poisson_predicted <- predict.glm(simple_poisson, type="response")

```


## Multivariate model 

### Poisson model 

```{r}
full_poisson <- glm(visits ~ age + income + illness +
                   reduced + health + private + freepoor                  + freerepat + nchronic + lchronic ,                         family="poisson",
                data = doc)

summary(full_poisson)
```
Have a look at the statistics of your model 

```{r}
glance(full_poisson)

```
Run an overdispersion test 

```{r}
dispersiontest(full_poisson, trafo = 1)
```

```{r}
dispersiontest(full_poisson, trafo = 2)
```

### Quasi-poisson

```{r}
full_quasipoisson <- glm(visits ~ age + income + illness +
                   reduced + health + private + freepoor                  + freerepat + nchronic + lchronic ,                         family="quasipoisson",
                data = doc)

summary(full_quasipoisson)
```
### Negative-binomial

```{r}
full_nb <- glm.nb(visits ~ age + income + illness +
                   reduced + health + private + freepoor                  + freerepat + nchronic + lchronic ,  
                data = doc)

summary(full_nb)
```

### Zero inflated poisson

```{r}
full_zeroinfl_poisson <-zeroinfl(visits ~ age + income +                                         illness +
                         reduced + health + private + freepoor                                   + freerepat + nchronic + lchronic ,  
                dist="poisson", data = doc)
summary(full_zeroinfl_poisson )
```

```{r}
logLik(full_poisson)
logLik(full_zeroinfl_poisson)
logLik(full_quasipoisson)
logLik(full_nb)
```

